---
title: "Lab 2: Vectors and other data structures"
author: "Amy Kim"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting the Data File

In this lab, you are going to work with a handful of variables about NBA players from the regular season 2016-2017:

* player: name of player.
* team: team name abbreviation.
* position: player position.
* age: age of player.
* experience: years of experience in NBA.
* salary: salary (in dollars).
* scored: total scored points.
* points1: number of free throws, worth 1 point each.
* points2: number of 2-point field goals, worth 2 points each.
* points3: number of 3-point field goals, worth 3 points each.

The data is in the file nba2017-salary-points.RData, located in the data/ folder of this github repository. The original source of the data is the website www.basketball-reference.com

Open a new session in Rstudio, and make sure you have a clean workspace by typing this command on the console:

```{r}
# remove existing objects
rm(list = ls())
```

You can download the .RData file to your working directory, and then load() it with the code below. Do NOT include these commands in your source Rmd file; simply type them directly on the console:

```{r}
# download RData file into your working directory
#rdata <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/data/nba2017-salary-points.RData"
#download.file(url = rdata, destfile = 'nba2017-salary-points.RData')
```

## What's happening in the code above?

The function download.file() allows you to download any type of file from the Web. In this case you are downloading the file called nba2017-salary-points.RData which is located in the github repository of the course. This file is a binary file. To be more precise, the file extension .RData is the default extension used by R for its binary native format.

Where does the file get downloaded? Bby default, the file nba2017-salary-points.RData gets downloaded to your working directory. If you are curious about what is the current directory to which R is paying attention to, simply type the function getwd()---which stands for get the working directory.

If you want to specify a specific location for the downloaded file, then modify the destfile parameter. For instance, if you are using Mac, and you want the file to be downloaded to your desktop, you can use:

```{r}
# download RData file to your Desktop (assuming you use Mac)
#rdata <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/data/nba2017-salary-points.RData"
#download.file(url = rdata, destfile = '~/Desktop/nba2017-salary-points.RData')
```

## Loading the data file
To load or import the contents of the binary file into your R session you use load(). This function allows you to import R binary files. This time, include the code below in your Rmd file:

```{r}
# load data in your R session
load('nba2017-salary-points.RData')
```

Note: the code above will only work as long as your Rmd file lives in the same directory of the .RData file. So far I'm assuming that you have both files in the working directory. If you run into problems, ask the GSI or any of the lab assistants.

Once you imported (or loaded) the data, use the function ls() which allows you to list all the available R objects:

```{r}
# list the available objects with ls()
ls()
```

Check that the following objects are available in your session:

* player
* team
* position
* salary
* etc

## About .RData files

Most of the data sets you are going to be working with in real life are going to be stored in some sort of text file. So you probably won't be handling many R binary files (i.e. .RData files). However, R binary files are an interesting option for saving intermediate results, and/or for saving R objects (as R objects).

## Inspecting the data objects

Once you have some data objects to work with, the first step is to inspect some of their characteristics. R has various functions that allow you to examine objects:

* typeof() type of storage of any object
* class() gives you the class of the object
* str() displays the structure of an object in a compact way
* mode() gives the data type (as used in R)
* object.size() gives an estimate of the memory space used by an object
* length() gives the length (i.e. number of elements)
* head() take a peek at the first elements
* tail() take a peek at the last elements
* summary() shows a summary of a given object


### Your turn:

* Use length(), head(), tail(), and summary() to start exploring the content of the loaded objects.

```{r}
length(player)
length(points)
length(position)
length(points)
length(points1)
length(points2)
length(points3)
length(salary)
length(team)
```


```{r}
head(player)
head(points)
head(position)
head(points)
head(points1)
head(points2)
head(points3)
head(salary)
head(team)
```


```{r}
tail(player)
tail(points)
tail(position)
tail(points)
tail(points1)
tail(points2)
tail(points3)
tail(salary)
tail(team)
```


```{r}
summary(player)
summary(points)
summary(position)
summary(points)
summary(points1)
summary(points2)
summary(points3)
summary(salary)
summary(team)
```

* Do all the objects have the same length?
Yes. Length of 441. 

* Are there missing values, i.e. NA, in any of the objects?

```{r}
sum(is.na(player))
sum(is.na(points))
sum(is.na(points1))
sum(is.na(points2))
sum(is.na(points3))
sum(is.na(position))
sum(is.na(salary))
sum(is.na(team))
```

There are no missing objects in the data

* Find out what is the class of each of the objects player, team, etc.

```{r}
class(player)
class(points)
class(points1)
class(points2)
class(points3)
class(position)
class(salary)
class(team)
```

* How do you know if any of the loaded objects is a vector?

```{r}
is.vector(player)
is.vector(points)
is.vector(points1)
is.vector(points2)
is.vector(points3)
is.vector(position)
is.vector(salary)
is.vector(team)
```

All of the objects are vectors expect for 'team'. 

* How do you know that a given vector is of a certain data type?
You can check with 'typeof()' command 

```{r}
typeof(salary)
typeof(position)
typeof(team)
```

# Manipulating Vectors: Subsetting

Create a vector four by selecting the first four elements in player:

```{r}
four <- head(player, n = 4)
```

Single brackets [ ] are used to subset (i.e. subscript, split) vectors. Find out what happens if you specify:

* number one: four[1]
```{r}
four[1]
```

* an index of zero: four[0]?
```{r}
four[0]
```

* a negative index: four[-1]?
```{r}
four[-1]
```

* various negative indices: four[-c(1,2,3)]?
```{r}
four[-c(1,2,3)]
```

* an index greater than the length of the vector: four[5]?
```{r}
four[5]
```

* repeated indices: four[c(1,2,2,3,3,3)]?
```{r}
four[c(1,2,2,3,3,3)]
```


Often, you will need to generate vectors of numeric sequences, like the first five elements 1:5, or from the first till the last element 1:length(player). R provides the colon operator :, and the functions seq(), and rep() to create various types of sequences.

Figure out how to use seq(), rep() to extract:

* all the even elements in player
```{r}
player[seq(2, length(player), by = 2)]
```

* all the odd elements in salary
```{r}
salary[seq(1, length(salary), by = 2) ]
```

* all multiples of 5 (e.g. 5, 10, 15, etc) of team
```{r}
team[seq(1, length(team), by = 5)]
```

* all the even elements in team but this time in reverse order
```{r}
sort(team[seq(1, length(team), by = 2)], decreasing = TRUE) 
```

Use logical subsetting occurs when the vector of indices that you pass inside the brackets is a logical vector.

Another kind of subsetting/subscripting is the so-called logical subsetting. This type of subsetting implies using a logical vector inside the brackets:

```{r}
four[c(TRUE, TRUE, TRUE, TRUE)]
four[c(TRUE, TRUE, FALSE, FALSE)]
four[c(FALSE, FALSE, TRUE, TRUE)]
four[c(TRUE, FALSE, TRUE, FALSE)]
four[c(FALSE, FALSE, FALSE, FALSE)]

# recycling
four[TRUE]
four[c(TRUE, FALSE)]
```

When subsetting a vector logically, most of the times you won't really be providing an explicit vector of TRUE's and FALSEs. Just imagine having a vector of 100 or 1000 or 1000000 elements, and trying to do logical subsetting by manually creating a logical vector of the same length. That would be very boring. Instead, you will be providing a logical condition or a comparison operation that returns a logical vector.

A comparison operation occurs when you use comparison operators such as:

* > greater than
* >= greater than or equal
* < less than
* <= less than or equal
* == equal
* != different

```{r}
a <- c(2, 4, 6, 7, 8, 10)

# elements greater than 6
a > 6

# elements different from 6
a != 6
```


Notice that a comparison operation always returns a logical vector. Here's how you actually extract the values of a vector based on logical indexing:

```{r}
points_five_hundred <- points[1:500]

# elements greater than 500
points_five_hundred[points_five_hundred > 500]

# elements less than 500
points_five_hundred[points_five_hundred < 500]

# elements less than or equal to 500
points_five_hundred[points_five_hundred <= 500]
```

In addition to using comparison operators, you can also use logical operators to produce a logical vector. The most common type of logical operators are:

* & AND
* | OR
* ! negation

Run the following commands to see what R does:

```{r}
# AND
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE

# OR
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE

# NOT
!TRUE
!FALSE
```

Logical operators allow you to combine several comparisons:

```{r}
# players of Golden State (GSW)
player[team == 'GSW']

# name of players with salaries greater than 20 million dollars
player[salary > 20000000]

# name of players with scored points between 1000 and 1200 (exclusive)
player[points > 1000 & points < 1200]
```

## Your Turn 

Write commands, using bracket notation, to answer the following questions (you may need to use min(), max(), which(), which.min(), which.max()):

* players in position Center, of Warriors (GSW)
```{r}
player[which(team == "GSW" & position == "C")]
```

* players of both GSW (warriors) and LAL (lakers)
```{r}
player[team == "GSW" | team == "LAL"]
```

* players in positions Shooting Guard and Point Guards, of Lakers (LAL)
```{r}
player[which(team == "LAL" & position == "SG" | team == "LAL" & position == "PG")]
```

* subset Small Forwards of GSW and LAL
```{r}
which(team == "GSW" & position == "SF" | team == "LAL" & position == "SF")

player[which(team == "GSW" & position == "SF" | team == "LAL" & position == "SF")]
```

* name of the player with largest salary
```{r}
player[which.max(salary)]
```

* name of the player with smallest salary
```{r}
player[which.min(salary)]
```

* name of the player with largest number of scored points
```{r}
player[which.max(points)]
```

* salary of the player with largest number of points
```{r}
salary[which.max(points)]
```

* largest salary of all Centers
```{r}
salary[which.max(salary[which(position == "C")])]
```

* team of the player with the largest number of scored points
```{r}
team[which.max(points)]
```

* name of the player with the largest number of 3-pointers
```{r}
player[which.max(points3)]
```


# Subsetting with Character Vectors

A third type of subsetting involves passing a character vector inside brackets. When you do this, the characters are supposed to be names of the manipulated vector.

None of the vectors player, position, and salary have names. You can confirm that with the names() function applied on any of the vectors:
```{r}
names(player)
```

* Create a vector warriors_player by selecting the players of team == 'GSW'
```{r}
warriors_player <- c(player[which(team == "GSW")])
```

* then create a vector warriors_salary with their salaries, and another vector warriors_points with their scored.
```{r}
warriors_salary <- c(salary[which(team == "GSW")])
warriors_points <- c(points[which(team == "GSW")])
```

* Assign warriors_player as the names of warriors_salary.
```{r}
names(warriors_salary) <- warriors_player
```

You should have a vector warriors_salary with named elements. Now you can use character subsetting:

```{r}
warriors_salary["Andre Iguodala"]

warriors_salary[c("Stephen Curry", "Kevin Durant")]

```

## Some Plotting 

Use the function plot() to make a scatterplot of points and salary

```{r}
plot(points, salary)
```

Keep in mind that plot() is a generic function. This means that the behavior of plot() depends on the type of input. When you pass two numeric vectors, plot() will attempt to create a scatter plot.

Looking at the generated plot, can you see any issues?

To get a better display of the scatterplot, let's create two vectors log_points and log_salary by transforming scored and salary with the logarithm function log()
```{r}
log_points <- log(points)
log_salary <- log(salary)
```

Make another scatterplot but now use the log-transformed vectors:

```{r}
plot(log_points, log_salary)

```

To add the names of the players in the plot, you can use the low-level graphing function text():

```{r}
plot(log_points, log_salary)
text(log_points, log_salary, labels = player)
```

Now we have another problem. The labels in the plot are very messy. A quick and dirty fix is to use abbreviate() to shorten the displayed names:

```{r}
plot(log_points, log_salary)
text(log_points, log_salary, labels = abbreviate(player))
```

## Your Turn: 

Create a scatterplot of points and salary for the Warriors (GSW), displaying the names of the players. Generate two scatterplots, one with raw values (original scale, and another plot with log-transformations).
```{r}
plot(warriors_points, warriors_salary)
text(warriors_points, warriors_salary, labels = abbreviate(warriors_player))
```

```{r}
log_warrior_points <- log(warriors_points)
log_warrior_salary <- log(warriors_salary)

plot(log_warrior_points, log_warrior_salary)
text(log_warrior_points, log_warrior_salary, labels = abbreviate(warriors_player))

```

# Vectorization

When you create the vectors log_scored <- log(scored) and log_salary <- log(salary), what you're doing is applying a function to a vector, which in turn acts on all elements of the vector.

This is called Vectorization in R parlance. Most functions that operate with vectors in R are vectorized functions. This means that an action is applied to all elements of the vector without the need to explicitly type commands to traverse all the elements.

In many other programming languages, you would have to use a set of commands to loop over each element of a vector (or list of numbers) to transform them. But not in R.

Another example of vectorization would be the calculation of the square root of all the elements in points and salary:

```{r}
sqrt(points)
sqrt(salary)

```

## Why should you care about vectorization?

If you are new to programming, learning about R's vectorization will be very natural (you won't stop to think about it too much). If you have some previous programming experience in other languages (e.g. C, python, perl), you know that vectorization does not tend to be a native thing.

Vectorization is essential in R. It saves you from typing many lines of code, and you will exploit vectorization with other useful functions known as the apply family functions (we'll talk about them later in the course).


# Recycling

Closely related with the concept of vectorization we have the notion of Recycling. To explain recycling let's see an example.

salary is given in dollars, but what if you need to obtain the salaries in millions of dollars?. Create a new vector salary_millions with the converted values in millions of dollars.

```{r}
salary_millions <- salary/1000000
```

Take the values in points1, points2, and points3 and figure out to create a new vector scored_points that gives you the same values of points.
```{r}
scored_points <- points1 + (2*points2) + (3*points3)
```

What you just did (assuming that you did things correctly) is called Recycling. To understand this concept, you need to remember that R does not have a data structure for scalars (single numbers). Scalars are in reality vectors of length 1.

# Factors

As mentioned before, vectors are the most essential type of data structure in R. They are atomic structures (can contain only one type of data): integers, real numbers, logical values, characters, complex numbers.

Related to vectors, there is another important data structure in R called factor. Factors are data structures exclusively designed to handle categorical data.

The object team is an R factor. You can confirm this by using is.factor() or class()

```{r}
is.factor(team)
```

## Creating Factors 

Use factor() to create an object position_fac by converting position into a factor:

```{r}
position_fac <- factor(position)
```

If you have a factor, you can invoke table() to get a table with the frequencies (i.e. counts) of the factor categories or levels:

```{r}
table(position_fac)
```


## Manipulating Factors

Because factors are internally stored as integers, you can manipulate factors as any other vector:

```{r}
position_fac[1:5]
```

Practice manipulating position_fac to get:

* positions of Warriors
```{r}
position_fac[team == "GSW"]
```

* positions of players with salaries > 15 millions
```{r}
position_fac[salary_millions > 15]
```

* frequencies (counts) of positions with salaries > 15 millions
```{r}
table(position_fac[salary_millions > 15])
```

* relative frequencies (proportions) of 'SG' (Shooting Guards) in each team
```{r}
table(team[position_fac == "SG"])
```


## More Plots

Let's go back to the scatterplot of points and salary

```{r}
plot(points, salary)
```

* Use your factor position_fac to add some color to the dots in the scatterplot.
```{r}
plot(points, salary)
points(points[position_fac == "C"], salary[position_fac == "C"], col = "red") # center
points(points[position_fac == "PF"], salary[position_fac == "PF"], col = "blue")
points(points[position_fac == "SG"], salary[position_fac == "SG"], col = "green")
points(points[position_fac == "PG"], salary[position_fac == "PG"], col = "purple")
```

* Pass the factor to the col = parameter inside plot()
```{r}
plot(points, salary, col = position_fac)
```

* Experiment with other plot() arguments like the point character pch =, the size of dots with the parameter cex =, the axes labels xlab and ylab and so on.
```{r}
plot(points, salary, col = position_fac, cex = 0.5 , xlab = "log of points of player", ylab = "log of salary of player", main = "Points to Salary ")

```




