---
title: "Lab 3: Command Line Basics and Git"
author: "Amy Kim"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic Bash shell commands

The first part of the lab involves navigating the file system and manipulating files (and directories) with the following basic bash commands:

* pwd: print working directory
* ls: list files and directories
* cd: change directory (move to another directory)
* mkdir: create a new directory
* touch: create a new (empty) file
* cp: copy file(s)
* mv: rename file(s)
* rm: delete file(s)

If you are using git-bash (i.e. your OS is Windows) you don't have the man command to see the manual documentation of other commands. In this case you can check the man pages online:

[link](http://man7.org/linux/man-pages/index.html)


## Your Turn 

Write your bash commands in a text editor (NOT a word processor) and save them in a text file: e.g. lab03-gaston-sanchez.txt.

* Open (or launch) the command line

* Use mkdir to create a new directory stat133-lab03

* Change to the directory stat133-lab03

* Use the command curl to download the following text file:
"curl -O http://textfiles.com/food/bread.txt"

* Use the command ls to list the contents in your current directory

* Use the command curl to download these other text files

[link](http://textfiles.com/food/btaco.txt)

[link](http://textfiles.com/food/1st_aid.txt)

[link](http://textfiles.com/food/beesherb.txt) 


* Use the command curl to download the following csv files:

[link](http://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/forestfires.csv) 

[link](http://www.math.uah.edu/stat/data/Fisher.csv)

[link](http://web.pdx.edu/~gerbing/data/cars.csv)

* Now try ls -l to list the contents in your current directory in long format

* Look at the man documentation of ls to find out how to list the contents in reverse order

* How would you list the contents in long format and by time?

* Inside stat133-lab03 create a directory data

* Change to the directory data

* Create a directory txt-files

* Create a directory csv-files

* Use the command mv to move the bread.txt file to the folder txt-files

* Use the wildcard * to move all the text files to the directory txt-files

* Use the wildcard * to move all the .csv files to the directory csv-files

* Go back to the parent directory stat133-lab03

* Create a directory copies

* Use the command cp to copy the bread.txt file (the one inside the folder txt-files) to the copies directory

* Use the wildcard * to copy all the .txt files in the directory copies

* Use the wildcard * to copy all the .csv files in the directory copies

* Change to the directory copies

* Use the command mv to rename the file bread.txt as bread-recipe.txt

* Rename the file Fisher.csv as iris.csv

* Rename the file btaco.txt as breakfast-taco.txt

* Change to the parent directory (i.e. stat133-lab03)

* Rename the directory copies as copy-files

* Find out how to use the rm command to delete the directory copy-files

* List the contents of the directory txt-files displaying the results in reverse (alphabetical) order


# Git and Github Practice

## 1) Create a New GitHub Repository

* Open your browser and Sign in to your github account.
* Locate the + button (next to your avatar).
* Select the New repository option.
* Choose a name for your repository: e.g. demo-repo.
* In the Description field add a brief description: e.g. "this is a demo repo"
* Use the default settings, and click the green button Create repository.
* You should see some content similar (but not identical) to the one below:

```{r}
#echo "# Demo Repo" >> README.md
#git init
#git add README.md
#git commit -m "first commit"
#git remote add origin https://github.com/gastonstat/demo-repo.git
#git push -u origin master
```

## 2) Create a local Git Repository

* Open the terminal (Mac Terminal, or Git-Bash for Windows users).

* Optional: change directory to your preferred location e.g. your Desktop

```{r}
#cd Desktop
```

* Create a directory with the name of your github repo
```{r}
#mkdir demo-repo
```

* Change to the directory you just created

```{r}
#cd demo-repo
```

* Initialize the directory as a git repository
```{r}
#git init
```

It's possible that you encounter some error message, e.g. Mac users may get a message related with a missing component for CommandLineTools. If this your case, then type in the terminal console:

```{r}
# Mac users may need to run this command
# xcode-select --install
```

The command git init will set-up your directory demo-repo as a Git repository (NOT to confuse with your GitHub repository). This is basically your local repository.



## 3) Adding a README file

It is customary to add a README.md file at the top level. This file must contain (at least) a description of what the repository is about. The following command will create a README.md file with some minimalist content:

```{r}
# echo "# Demo Repo" >> README.md
```

* So far there you have a "new" file in your local repo, but this change has not been recorded by Git. You can confirm this by checking the status of the repo:

```{r}
# git status
```

* Notice that Git knows that README.md is untracked. So let's add the changes to Git's database:
```{r}
# git add README.md
```

* Check the status of the repo again:

```{r}
#git status
```

* Now Git is tracking the file README.md.

* Next thing consists of committing the changes

```{r}
#git commit -m "first commit"
```


## 4) Adding a remote

Right now you have a (local) Git repository in your computer. And you also have a GitHub repository in your GitHub account. Both repositories should have the same name, and the goal is to link them. To do this, you need to tell Git that a remote repository (i.e. the one in GitHub) will be added:

To add a remote repository use the command below with your own username:

```{r}
# git remote add origin https://github.com/username/demo-repo.git
```

Verify your new remote

```{r}
# git remote -v
```

If everything is okay, you should be able to see a message (with your own username) like this:
```{r}

# Verify new remote
#origin  https://github.com/username/demo-repo.git (fetch)
#origin  https://github.com/username/demo-repo.git (push)

```

## Pushing changes to a remote repo

* Now that you have linked your local repo with your remote repo, you can start pushing (i.e. uploading) commits to GitHub.

* As part of the basic workflow with git and github, you want to constantly check the status of your repo

```{r}
#git status
```

* Now let's push your recent commit to the remote branch (origin) from the local branch (master):

```{r}
#git push origin master
```


* Go to your Github repository and refresh the browser. If everything went fine, you should be able to see the contents of your customized README.md file.




