---
title: 'Lab 6: More data wrangling and outputs'
author: "Amy Kim"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Manipulating and Visualizing Data Frames

In this lab, you will continue manipulating data frames with "dplyr", and plotting graphics with "ggplot2". In addition, you will also use various functions to export (or save) tables, images, and R output to external files.


## Filestructure

To help you better prepare for HW02, we want you to practice working with a more sophisticated file structure. Follow the steps listed below to create the necessary subdirectories like those depicted in this scheme:
```{r}
#lab06/
#       README.md
#       data/
#       code/
#       output/
#       images/
```

* Open a shell terminal (e.g. command line or GitBash)

* Change your working directory to a location where you will store all the materials for this lab

* Use mkdir to create a directory lab06

* cd to lab06

* Use mkdir to create other subdirectories: data, code, output, images

* List the contents of lab06 to confirm that you have all the subdirectories

* Use touch to create an empty README.md text file.

* Open the README.md file with a text editor (e.g. the one in RStudio) and add a brief description of what this lab is about. Save the changes.

* cd to data/

* Download the data file with the command curl, and the -O option (letter O)

```{r}
# curl -O https://raw.githubusercontent.com/ucb-stat133/stat133-spring-2018/master/data/nba2017-players.csv
```

* Use ls to confirm that the csv file is in data/

* Use word count wc to count the lines of the csv file
```{r}
#      442    1632   39752 nba2017-players.csv
```

* Take a peek at the first rows of the csv file with head
```{r}
#"player","team","position","height","weight","age","experience","college","salary","games","minutes","points","points3","points2","points1"
#"Al Horford","BOS","C",82,245,30,9,"University of Florida",26540100,68,2193,952,86,293,108
#"Amir Johnson","BOS","PF",81,240,29,11,"",1.2e+07,80,1608,520,27,186,67
#"Avery Bradley","BOS","SG",74,180,26,6,"University of Texas at #Austin",8269663,55,1835,894,108,251,68
#"Demetrius Jackson","BOS","PG",73,201,22,0,"University of Notre Dame",1450000,5,17,10,1,2,3
#"Gerald Green","BOS","SF",79,205,31,9,"",1410598,47,538,262,39,56,33
#"Isaiah Thomas","BOS","PG",69,185,27,5,"University of Washington",6587132,76,2569,2199,245,437,590
#"Jae Crowder","BOS","SF",78,235,26,4,"Marquette University",6286408,72,2335,999,157,176,176
#"James Young","BOS","SG",78,215,21,2,"University of Kentucky",1825200,29,220,68,12,13,6
#"Jaylen Brown","BOS","SF",79,225,20,0,"University of California",4743000,78,1341,515,46,146,85

```

* Take a peek at the last 5 rows of the csv file with tail
```{r}
#"Dragan Bender","PHO","PF",85,225,19,0,"",4276320,43,574,146,28,29,4
#"Elijah Millsap","PHO","SG",78,225,29,2,"University of Alabama at Birmingham",23069,2,23,3,0,1,1
#"Eric Bledsoe","PHO","PG",73,190,27,6,"University of Kentucky",1.4e+07,66,2176,1390,104,345,388
#"Jared Dudley","PHO","PF",79,225,31,9,"Boston College",10470000,64,1362,434,77,80,43
#"Leandro Barbosa","PHO","SG",75,194,34,13,"",4e+06,67,963,419,35,137,40
#"Marquese Chriss","PHO","PF",82,233,19,0,"University of Washington",2941440,82,1743,753,72,212,113
#"Ronnie Price","PHO","PG",74,190,33,11,"Utah Valley State College",282595,14,134,14,3,1,3
#"T.J. Warren","PHO","SF",80,230,23,2,"North Carolina State University",2128920,66,2048,951,26,377,119
#"Tyler Ulis","PHO","PG",70,150,21,0,"University of Kentucky",918369,61,1123,444,21,163,55
#"Tyson Chandler","PHO","C",85,240,34,15,"",12415000,47,1298,397,0,153,91
```

# R script file

* Once you have the filestructure for this lab, go to RStudio and open a new R script file (do NOT confuse with an Rmd file).

* Save the R script file as lab06-script.R in the code/ folder of lab06/

* R script files are used to write R code only, using R syntax. In other words, you should NOT use Markdown or LaTeX syntax inside an R script file. Why? Because if you run the entire script, R will try to execute all the commands, and won't be able to recognize Markdown, LaTeX, yaml, or other syntaxes.

## File Header

Let's start with some good coding practices by adding a header to the R script file in the form of R comments. In general, the header section should contain a title, a description of what the script is about, what are the inputs, and what are the main outputs produced when executing the code in the script. Optionally, you can also include the name of the author, the date, and other details. Something like this:

```{r}
# Title: Short title (one sentence)
# Description: what the script is about (one paragraph or two)
# Input(s): what are the main inputs (list of inputs)
# Output(s): what are the main outputs (list of outputs)
# Author(s): First Last
# Date: mm-dd-yyyy

```

Think of the header of a script file as the yaml header used in Rmd files. The header should be the very first thing that appears at the top of the script file. Personally, I like to surround the header in my R script files with some delimiting characters that help the reader to visually identify main parts of the script. Here's a hypothetical example of a header:

```{r}
# ===================================================================
# Title: Cleaning Data
# Description:
#   More data wrangling and outputs on csv file
# Input(s): data file 'raw-data.csv'
# Output(s): data file 'clean-data.csv'
# Author: Gaston Sanchez
# Date: 2-27-2018
# ===================================================================
```

Another good coding practice is to avoid writing very long lines of code. Most coding style guides stick to a maximum line width of 80 characters, and this is the magic number that I also use for my scripts.

**Your turn**: Include a header in your R script file, respecting the width limit of 80 characters. Save this file in the code/ folder.

```{r}
# ===================================================================
# Title: lab 6 
# Description:
#   This script performs cleaning tasks and transformations on 
#   various columns of the raw data file.
# Input(s): data file 'nba2017-players.csv'
# Output(s): data file 'lab6-scrip.R' and ggplot
# Author: Amy Kim 
# Date: 2-27-2018
# ===================================================================

```



## Required Packages

The next thing that you need to include in your script file are the required packages. Not all script files need packages, but many do. When this is the case, loading the packages should be the first lines of code to be executed.

Include the commands to load the following packages in your script:
```{r}
# packages
library(readr)    # importing data
library(dplyr)    # data wrangling
library(ggplot2)  # graphics
```

In addition to loading the packages, sometimes you will also need to load code from other script files. We won't do that today, but you should know that this is very common as the complexity and size of your projects grow.

## Exporting some data tables

After the header, and the loading-packages sections, the next part in your lab script involves importing the data. In addition to importing a data table, you are also going to practice exporting tables. That is, writing data tables to external files.

* We want you to work with relative paths. To execute the commands that read input(s) and write output(s), you will need to set the working directory. On the menu bar go to Session, then select the option Set Working Directory, and choose To Source File Location.

* Use read_csv() from the package "readr" to import the data nba2017-players.csv in R. Do this by specifying a relative path.
```{r}
dat <- read_csv('/Users/akii_my/Desktop/school/stat133/labs/lab6/data/nba2017-players.csv')
```

* Use the imported tibble to create a data frame warriors by selecting rows---e.g. filter()---of Golden State Warriors, arranging rows by salary in increasing order.
```{r}
gsw <- dat %>%
  filter(team == "GSW") %>%
  arrange(salary)
```

* Use the function write.csv() to export (or save) the data frame warriors to a data file warriors.csv in the folder/ directory. You will need to use a relative path to specify the file argument. Also, see how to use the argument row.names to avoid including a first column of numbers.
```{r}
write.csv(gsw, file = '/Users/akii_my/Desktop/school/stat133/labs/lab6/data/warriors.csv', row.names = FALSE)
```

* Create another data frame lakers by selecting rows of Los Angeles Lakers, this time arranging rows by experience (decreasingly).
```{r}
lakers <- dat %>% 
  filter(team == "Los Angeles Lakers") %>%
  arrange(desc(experience))
```


* Now use the function write_csv() to export (or save) the data frame lakers to a data file lakers.csv in the folder/ directory. You will also need to use a relative path to specify the file argument.
```{r}
write.csv(gsw, file = '/Users/akii_my/Desktop/school/stat133/labs/lab6/data/lakers.csv', row.names = FALSE)
```

* Inspect the contents of the data/ folder and confirm that the csv files are there.

## Exporting some R output

After exporting the tables to the corresponding csv files, you will produce some summary statistics, and then save the generated output to external text files. To do this, you will have to learn about the sink() function, which sends R output to a specified file.

Say you are interested in exporting the summary statistics of height and weight, exactly in the same way they are displayed by R:

```{r}
summary(dat[ ,c('height', 'weight')])
```

One naive option to "export" this output would be to manually copy the text displayed on the console, and then paste it to a text file. While this may work, it is labor intensive, error prone, and highly irreproducible. A better way to achieve this task is with the sink() function. Here's how:

```{r}
# divert output to the specified file
sink(file = '/Users/akii_my/Desktop/school/stat133/labs/lab6/output/summary-height-weight.txt')
summary(dat[ ,c('height', 'weight')])
sink()
```

The fist call to sink() opens a connection to the specified file, and then all outputs are diverted to that location. The second call to sink(), i.e. the one without any arguments, closes the connection.

While you are sink()ing output to a specified file, all the results will be sent to such file. In other words, nothing will be printed on the console. Only after the sinking process has finished and the connection is closed, you will be able to execute commands and see results displayed on R's console.

### Why sinking?

Why would you ever want to sink() R outputs to a file? Why not simply display them as part of your Rmd file? One good reason for diverting output to an external file is for convenience. In practice, the reports and documents (e.g. papers, executive summaries, slides) of a data analysis project won't contain everything that you tried, explored, analyzed, and graphed. There will be many intermediate results that, while relevant for a specific stage of the analysis cycle, are innecessary for the final report. So a good way to keep these intermediate outputs is by exporting them with sink().

#### Your turn:

* Export the output of str() on the data frame with all the players. sink() the output, using a relative path, to a text file data-structure.txt, in the output/ folder.
```{r}
sink(file = '/Users/akii_my/Desktop/school/stat133/labs/lab6/output/data-structure.txt')
str(dat)
sink()
```

* Export the summary() of the entire data frame warriors to a text file summary-warriors.txt, in the output/ folder (also use a relative path).
```{r}
sink(file = '/Users/akii_my/Desktop/school/stat133/labs/lab6/output/summary-warriors.txt')
summary(gsw)
sink()
```

* Export another summary() of the entire data frame lakers to a text file summary-lakers.txt, in the output/ folder (using a relative path).
```{r}
sink(file = '/Users/akii_my/Desktop/school/stat133/labs/lab6/output/summary-lakers.txt')
summary(lakers)
sink()
```


## Exporting some "base" graphs

n the same way that R output, as it appears on the console, can be exported to some files, you can do the same with graphics and plots. Actually, saving plot images is much more common than sink()ing output.

Base R provides a wide array of functions to save images in most common formats:

* png()
* jpeg()
* tiff()
* bmp()
* svg()
* pdf()

Similar to the writing table functions such as write.table() or write.csv(), and the sink() function, the graphics device functions require a file name to be provided. Here's how to save a simple scatterpot of height and weight in png format to the folder images/:

```{r}
# saving a scatterplot in png format
png(filename = "/Users/akii_my/Desktop/school/stat133/labs/lab6/images/scatterplot-height-weight.png")
plot( y = dat$height, x = dat$weight, pch = 20, 
     xlab = 'Weight', ylab = 'Height')
dev.off()
```

* The function png() tells R to save the image in PNG format, using the provided filename.
* Invoking png() will open a graphics device; not the graphics device of RStudio, so you won't be able to see the graphic.
* The plot() function produces the scatterplot.
* The function dev.off() closes the graphics device.

#### Your Turn: 

* Open the help documentation of png() and related graphic devices.

* Use png() to save a scatterplot of height and weight with plot(). Save the graph in the images/ folder.
```{r}
png(filename = "/Users/akii_my/Desktop/school/stat133/labs/lab6/images/scatterplot-height-weight.png")
plot( y = dat$height, x = dat$weight, pch = 20, 
     xlab = 'Weight', ylab = 'Height')
dev.off()
```

* Save another version of the scatterplot between height and weight, but now try to get an image with higher resolution. Save the plot in images/.
```{r}
png(filename = "/Users/akii_my/Desktop/school/stat133/labs/lab6/images/high-scatterplot-height-weight.png", res = 110)
plot( y = dat$height, x = dat$weight, pch = 20, 
     xlab = 'Weight', ylab = 'Height')
dev.off()
```

* Save a histogram in JPEG format of age with dimensions (width x height) 600 x 400 pixels. Save the plot in images/.
```{r}
jpeg(filename = "/Users/akii_my/Desktop/school/stat133/labs/lab6/images/scatterplot-height-weight.jpeg", width = 600, height = 400, units = "px")
plot( y = dat$height, x = dat$weight, pch = 20, 
     xlab = 'Weight', ylab = 'Height')
dev.off()
```

* Use pdf() to save the previous histogram of age in PDF format, with dimensions (width x height) 7 x 5 inches.
```{r}
pdf(file = "/Users/akii_my/Desktop/school/stat133/labs/lab6/images/scatterplot-height-weight.pdf", width = 7, height = 5)
plot(y = dat$height, x = dat$weight, pch = 20, 
     xlab = 'Weight', ylab = 'Height')
dev.off()
```


## Exporting some ggplots


The package "ggplot2" comes with a wrapper function ggsave() that allows you to save ggplot graphics to a specified file. By default, ggsave() saves images in PDF format.

* Use ggplot() to make a scatterplot of points and salary, and store it in a ggplot object named gg_pts_salary. Then use ggsave() to save the plot with dimensions (width x height) 7 x 5 inches; in the images/ folder as points_salary.pdf
```{r}
gg_pts_salary <- ggplot(dat, aes(x= points, y = salary)) + 
  geom_point()

ggsave(
  filename = "/Users/akii_my/Desktop/school/stat133/labs/lab6/images/points_salary.pdf",
  plot = last_plot(),
  device = "pdf",
  path = NULL,
  scale = 1,
  width = 7,
  height = 5,
  units = "in"
)

```

* Use ggplot() to create a scatterplot of height and weight, faceting by position. Store this in a ggplot object gg_ht_wt_positions Then use ggsave() to save the plot with dimensions (width x height) 6 x 4 inches; in the images/ folder as height_weight_by_position.pdf
```{r}
height_weight_by_position <- ggplot(dat, aes(x= weight, y = height)) + 
  geom_point() + 
  facet_grid(~ position)

ggsave(
  filename = "/Users/akii_my/Desktop/school/stat133/labs/lab6/images/height_weight_by_position.pdf",
  plot = last_plot(),
  device = "pdf",
  path = NULL,
  scale = 1,
  width = 6,
  height = 4,
  units = "in"
)

```


# More "dplyr"

The last part of this lab involves working the pipe operator %>% which allows you write function calls in a more human-readable way. This becomes extremely useful in "dplyr" operations that require many steps.

The behavior of "dplyr" is functional in the sense that function calls don't have side-effects. You must always save their results in order to keep them in an object (in memory). This doesn't lead to particularly elegant code, especially if you want to do many operations at once. You either have to do it step-by-step:

```{r}
# manipulation step-by-step
dat1 <- group_by(dat, team)
dat2 <- select(dat1, team, height, weight)
dat3 <- summarise(dat2,
  avg_height = mean(height, na.rm = TRUE),
  avg_weight = mean(weight, na.rm = TRUE))
dat4 <- arrange(dat3, avg_height)
dat4

```

Or if you don't want to name the intermediate results, you need to wrap the function calls inside each other:

```{r}
# inside-out style (hard to read)
arrange(
  summarise(
    select(
      group_by(dat, team),
      team, height, weight
    ),
    avg_height = mean(height, na.rm = TRUE),
    avg_weight = mean(weight, na.rm = TRUE)    
  ),
  avg_height
)

```


This is difficult to read because the order of the operations is from inside to out. Thus, the arguments are a long way away from the function. To get around this problem, "dplyr" provides the %>% operator from "magrittr".

x %>% f(y) turns into f(x, y) so you can use it to rewrite multiple operations that you can read left-to-right, top-to-bottom:

```{r}
# using %>%
dat %>% 
  group_by(team) %>%
  select(team, height, weight) %>%
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_weight = mean(weight, na.rm = TRUE)) %>%
  arrange(avg_height)
```

Use the piper operator "%>%" to perform the following operations:

* display the player names of Lakers 'LAL'.
```{r}
dat %>%
  filter(team == "LAL") %>%
  select(player)
```

* display the name and salary of GSW point guards 'PG'.
```{r}
dat %>%
  filter(team == "GSW" & position == "PG") %>%
  select(player, salary)
```

* dislay the name, age, and team, of players with more than 10 years of experience, making 10 million dollars or less.
```{r}
dat %>%
  filter(experience > 10 & salary < 1000000 | salary == 1000000) %>%
  select(player, age, team)
```

* select the name, team, height, and weight, of rookie players, 20 years old, displaying only the first five occurrences (i.e. rows).
```{r}
dat %>%
  filter(experience < 1 & age == 20) %>%
  select(player, team, height, weight) %>%
  slice(1:5)
```

* create a data frame gsw_mpg of GSW players, that contains variables for player name, experience, and min_per_game (minutes per game), sorted by min_per_game (in descending order).
```{r}
dat %>%
  filter(team == "GSW") %>%
  summarise(player, experience, min_per_game = games/minutes) %>%
  arrange(desc(min_per_game))
```

* display the average triple points by team, in ascending order, of the bottom-5 teams (worst 3pointer teams).
```{r}
dat %>%
  group_by(team) %>%
  summarise(ave_triple = mean(points3)) %>%
  arrange(ave_triple) %>%
  slice(5:10)
```

* obtain the mean and standard deviation of age, for Power Forwards, with 5 and 10 years (including) of experience.
```{r}
dat %>%
  filter(position == "PF" & experience < 10 | experience > 5 | experience  == 10 | experience == 5) %>%
  summarise(mean_age = mean(age), sd_age = sd(age))
```
