---
title: 'Lab 5: First contact with dplyr and ggplot2'
author: "Amy Kim"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.path = '/Users/akii_my/Desktop/school/stat133/labs/lab5/images/')
```

## Manipulating and Visualizing Data Frames

Last week you started to manipulate data tables (under the class of "data.frame" objects) using bracket notation, dat[ , ], and the dollar operator, dat$name, in order to select specific rows, columns, or cells. In addition, you have been creating charts with functions like plot(), boxplot(), and barplot(), which are part of the "graphics" package.

In this lab, you will start learning about other approaches to manipulate tables and create statistical charts. We are going to use the functionality of the package "dplyr" to work with tabular data in a more consistent way. This is a fairly recent package introduced a couple of years ago, but it is based on more than a decade of research and work lead by Hadley Wickham.

Likewise, to create graphics in a more consistent and visually pleasing way, we are going to use the package "ggplot2", also originally authored by Hadley Wickham, and developed as part of his PhD more than a decade ago.

Use the first hour of the lab to get as far as possible with the material associated to "dplyr". Then use the second hour of the lab to work on graphics with "ggplot2".

## Filestructure and Shell Commands

We want you to keep practicing with the command line (e.g. Mac Terminal, Gitbash). Follow the steps listed below to create the necessary subdirectories like those depicted in this scheme:

```{r}
#    lab05/
#      README.md
#      data/
#        nba2017-players.csv
#      report/
#        lab05.Rmd
#        lab05.html
#      images/
#        ... # all the plot files
```

* Open a command line interface (e.g. Terminal or GitBash)

* Change your working directory to a location where you will store all the materials for this lab

* Use mkdir to create a directory lab05 for the lab materials

* Use cd to change directory to (i.e. move inside) lab05

* Create other subdirectories: data, report, images

* Use ls to list the contents of lab05 and confirm that you have all the subdirectories.

* Use touch to create an empty README.md text file

* Use a text editor (e.g. the one in RStudio) to open the README.md file, and then add a brief description of today's lab, using markdown syntax.

* Change directory to the data/ folder.

* Download the data file with the command curl, and the -O option (letter O)

```{r}
# curl -O https://raw.githubusercontent.com/ucb-stat133/stat133-spring-2018/master/data/nba2017-players.csv
```

* Use ls to confirm that the csv file is in data/

* Use word count wc to count the lines of the csv file
```{r}
#wc nba2017-players.csv
#  442    1632   39752 nba2017-players.csv
```

* Take a peek at the first rows of the csv file with head
```{r}
#"player","team","position","height","weight","age","experience","college","salary","games","minutes","points","points3","points2","points1"
#"Al Horford","BOS","C",82,245,30,9,"University of Florida",26540100,68,2193,952,86,293,108
#"Amir Johnson","BOS","PF",81,240,29,11,"",1.2e+07,80,1608,520,27,186,67
#"Avery Bradley","BOS","SG",74,180,26,6,"University of Texas at Austin",8269663,55,1835,894,108,251,68
#"Demetrius Jackson","BOS","PG",73,201,22,0,"University of Notre Dame",1450000,5,17,10,1,2,3
#"Gerald Green","BOS","SF",79,205,31,9,"",1410598,47,538,262,39,56,33
#"Isaiah Thomas","BOS","PG",69,185,27,5,"University of Washington",6587132,76,2569,2199,245,437,590
#"Jae Crowder","BOS","SF",78,235,26,4,"Marquette University",6286408,72,2335,999,157,176,176
#"James Young","BOS","SG",78,215,21,2,"University of Kentucky",1825200,29,220,68,12,13,6
#"Jaylen Brown","BOS","SF",79,225,20,0,"University of California",4743000,78,1341,515,46,146,85
```

Take a peek at the last 5 rows of the csv file with tail
```{r}
#"Dragan Bender","PHO","PF",85,225,19,0,"",4276320,43,574,146,28,29,4
#"Elijah Millsap","PHO","SG",78,225,29,2,"University of Alabama at Birmingham",23069,2,23,3,0,1,1
#"Eric Bledsoe","PHO","PG",73,190,27,6,"University of Kentucky",1.4e+07,66,2176,1390,104,345,388
#"Jared Dudley","PHO","PF",79,225,31,9,"Boston College",10470000,64,1362,434,77,80,43
#"Leandro Barbosa","PHO","SG",75,194,34,13,"",4e+06,67,963,419,35,137,40
#"Marquese Chriss","PHO","PF",82,233,19,0,"University of Washington",2941440,82,1743,753,72,212,113
#"Ronnie Price","PHO","PG",74,190,33,11,"Utah Valley State College",282595,14,134,14,3,1,3
#"T.J. Warren","PHO","SF",80,230,23,2,"North Carolina State University",2128920,66,2048,951,26,377,119
#"Tyler Ulis","PHO","PG",70,150,21,0,"University of Kentucky",918369,61,1123,444,21,163,55
#"Tyson Chandler","PHO","C",85,240,34,15,"",12415000,47,1298,397,0,153,91
```

## Installing packages

I'm assuming that you already installed the packages "dplyr" and "ggplot2". If that's not the case then run on the console the command below (do NOT include this command in your Rmd)

Remember that you only need to install a package once! After a package has been installed in your machine, there is no need to call install.packages() again on the same package. What you should always invoke in order to use the functions in a package is the library() function:

```{r}
library(dplyr)
library(ggplot2)
```

**About loading packages**: Another rule to keep in mind is to always load any required packages at the very top of your script files (.R or .Rmd or .Rnw files). Avoid calling the library() function in the middle of a script. Instead, load all the packages before anything else.

## Path for Images

The other important specification to include in your Rmd file is a global chunk option to specify the location of plots and graphics. This is done by setting the fig.path argument inside the knitr::opts_chunk$set() function.

If you don't specify fig.path, "knitr" will create a default directory to store all the plots produced when knitting an Rmd file. This time, however, we want to have more control over where things are placed. Because you already have a folder images/ as part of the filestructure, this is where we want "knitr" to save all the generated graphics.

Notice the use of a relative path fig.path = '../images/'. This is because your Rmd file should be inside the folder report/, but the folder images/ is outside report/ (i.e. in the same parent directory of report/).

# NBA Players Data

The data file for this lab is the same you used last week: nba2017-players.csv.

To import the data in R you can use the base function read.csv(), or you can also use read_csv() from the package "readr":

```{r}
# with "base" read.csv()
dat <- read.csv('~/Desktop/school/stat133/labs/lab5/data/nba2017-players.csv', stringsAsFactors = FALSE)

library(readr)
# with "readr" read_csv()
dat <- read_csv('~/Desktop/school/stat133/labs/lab5/data/nba2017-players.csv')
```

# Basic "dplyr" verbs

To make the learning process of "dplyr" gentler, Hadley Wickham proposes beginning with a set of five basic verbs or operations for data frames (each verb corresponds to a function in "dplyr"):

* filter: keep rows matching criteria
* select: pick columns by name
* mutate: add new variables
* arrange: reorder rows
* summarise: reduce variables to values

I've slightly modified Hadley's list of verbs:

* filter(), slice(), and select(): subsetting and selecting rows and columns
* mutate(): add new variables
* arrange(): reorder rows
* summarise(): reduce variables to values
* group_by(): grouped (aggregate) operations

# Filtering, slicing, and selecting

slice() allows you to select rows by position:

```{r}
# first three rows
three_rows <- slice(dat, 1:3)
three_rows
```

filter() allows you to select rows by condition:

```{r}
# subset rows given a condition
# (height greater than 85 inches)
gt_85 <- filter(dat, height > 85)
gt_85
```

select() allows you to select columns by name:

```{r}
# columns by name
player_height <- select(dat, player, height)
```

## Your turn:

* use slice() to subset the data by selecting the first 5 rows.
```{r}
slice(dat, n = 1:5)
```

* use slice() to subset the data by selecting rows 10, 15, 20, ..., 50.
```{r}
slice(dat, n = 10, 15, 20, 25, 30, 35, 40, 45, 50)
```

* use slice() to subset the data by selecting the last 5 rows.
```{r}
slice(dat, n = 436:441)
```

* use filter() to subset those players with height less than 70 inches tall.
```{r}
dat %>%
  filter(height  < 70)
```

* use filter() to subset rows of Golden State Warriors ('GSW').
```{r}
dat %>%
  filter(team  == "GSW")
```

* use filter() to subset rows of GSW centers ('C').
```{r}
dat %>%
  filter(team  == "GSW" & position == "C")
```

* use filter() and then select(), to subset rows of lakers ('LAL'), and then display their names.
```{r}
dat %>% 
  filter(team == "LAL") %>%
  select(player)
```

* use filter() and then select(), to display the name and salary, of GSW point guards
```{r}
dat %>% 
  filter(team == "GSW" & position == "PG") %>%
  select(player, salary)
```

* find how to select the name, age, and team, of players with more than 10 years of experience, making 10 million dollars or less.
```{r}
dat %>%
  filter(experience > 10 & salary < 10000000) %>%
  select(player, age, team)
```

* find how to select the name, team, height, and weight, of rookie players, 20 years old, displaying only the first five occurrences (i.e. rows)
```{r}
dat %>%
  filter(experience < 1 & age == 20) %>%
  select(player, team, height, weight) %>%
  slice_tail(n = 5)
```

# Adding new variables: mutate()

Another basic verb is mutate() which allows you to add new variables. Let's create a small data frame for the warriors with three columns: player, height, and weight:

```{r}
# creating a small data frame step by step
gsw <- filter(dat, team == 'GSW')
gsw <- select(gsw, player, height, weight)
gsw <- slice(gsw, c(4, 8, 10, 14, 15))
gsw
```

Now, let's use mutate() to (temporarily) add a column with the ratio height / weight:

```{r}
mutate(gsw, height / weight)
```

You can also give a new name, like: ht_wt = height / weight:

```{r}
mutate(gsw, ht_wt = height / weight)
```

In order to permanently change the data, you need to assign the changes to an object:
```{r}
gsw2 <- mutate(gsw, ht_m = height * 0.0254, wt_kg = weight * 0.4536)
gsw2
```

# Reordering rows: arrange()

The next basic verb of "dplyr" is arrange() which allows you to reorder rows. For example, here's how to arrange the rows of gsw by height

```{r}
# order rows by height (increasingly)
arrange(gsw, height)
```

By default arrange() sorts rows in increasing order. To arrange rows in descending order you need to use the auxiliary function desc().

```{r}
# order rows by height (decreasingly)
arrange(gsw, desc(height))

# order rows by height, and then weight
arrange(gsw, height, weight)
```

## Your Turn

* using the data frame gsw, add a new variable product with the product of height and weight.
```{r}
mutate(gsw, h_w = height * weight)
```

* create a new data frame gsw3, by adding columns log_height and log_weight with the log transformations of height and weight.
```{r}
gsw3 <- mutate(gsw, log_height = log(height), log_weight = log(weight))
gsw3
```

* use the original data frame to filter() and arrange() those players with height less than 71 inches tall, in increasing order.
```{r}
dat %>%
  filter(height < 71) %>%
  arrange(height)
```

* display the name, team, and salary, of the top-5 highest paid players
```{r}
dat %>%
  arrange(desc(salary)) %>%
  slice(n = 1:5) %>%
  select(player, team, salary)
```

* display the name, team, and points3, of the top 10 three-point players
```{r}
dat %>%
  arrange(desc(points3)) %>%
  slice(n = 1:10) %>%
  select(player, team, points3)
```

* create a data frame gsw_mpg of GSW players, that contains variables for player name, experience, and min_per_game (minutes per game), sorted by min_per_game (in descending order)
```{r}

gsw_mpg <- dat %>%
  filter(team == "GSW") %>%
  mutate(min_per_game = minutes / games) %>%
  arrange(desc(min_per_game)) %>%
  select(player, experience, min_per_game)
gsw_mpg
```

# Summarizing values with summarise()

The next verb is summarise(). Conceptually, this involves applying a function on one or more columns, in order to summarize values. This is probably easier to understand with one example.

Say you are interested in calculating the average salary of all NBA players. To do this "a la dplyr" you use summarise(), or its synonym function summarize():

```{r}
# average salary of NBA players
summarise(dat, avg_salary = mean(salary))
```

Calculating an average like this seems a bit verbose, especially when you can directly use mean() like this:

```{r}
mean(dat$salary)
```

So let's make things a bit more interessting. What if you want to calculate some summary statistics for salary: min, median, mean, and max?

```{r}
# some stats for salary (dplyr)
summarise(
  dat, 
  min = min(salary),
  median = median(salary),
  avg = mean(salary),
  max = max(salary)
)
```

Well, this may still look like not much. You can do the same in base R (there are actually better ways to do this):

```{r}
# some stats for salary (base R)
c(min = min(dat$salary), 
  median = median(dat$salary),
  median = mean(dat$salary),
  max = max(dat$salary))
```

# Grouped operations

To actually appreciate the power of summarise(), we need to introduce the other major basic verb in "dplyr": group_by(). This is the function that allows you to perform data aggregations, or grouped operations.

Let's see the combination of summarise() and group_by() to calculate the average salary by team:

```{r}
# average salary, grouped by team
summarise(
  group_by(dat, team),
  avg_salary = mean(salary)
)
```

Here's a similar example with the average salary by position:

```{r}
# average salary, grouped by position
summarise(
  group_by(dat, position),
  avg_salary = mean(salary)
)

```

Here's a more fancy example: average weight and height, by position, displayed in desceding order by average height:

```{r}
arrange(
  summarise(
    group_by(dat, position),
    avg_height = mean(height),
    avg_weight = mean(weight)),
  desc(avg_height)
)

```

## Your turn:

* use summarise() to get the largest height value.
```{r}
summarise(
  group_by(dat),
  max_height = max(height)
)
```


* use summarise() to get the standard deviation of points3.
```{r}
summarise(
  group_by(dat),
  sd_points3 = sd(points3) 
)
```

* use summarise() and group_by() to display the median of three-points, by team.
```{r}
summarise(
  group_by(dat, team),
  median_points3 = median(points3)
)
```

* display the average triple points by team, in ascending order, of the bottom-5 teams (worst 3pointer teams)
```{r}
slice_tail(
  arrange(
    summarise(group_by(dat, team),mean_points3 = mean(points3))
    , mean_points3)
  , n = 5)
```

* obtain the mean and standard deviation of age, for Power Forwards, with 5 and 10 years (including) years of experience.
```{r}

dat %>%
  filter(position == "PF" & experience > 5 & experience < 10 | 
           experience == 5 | experience == 10) %>%
  summarise(mean_age = mean(age), sd_age = sd(age))

```


# First contact with ggplot()

The package "ggplot2" is probably the most popular package in R to create beautiful static graphics. Comapred to the functions in the base package "graphcics", the package "ggplot2" follows a somewhat different philosophy, and it tries to be more consistent and modular as possible.

* The main function in "ggplot2" is ggplot()
* The main input to ggplot() is a data frame object.
* You can use the internal function aes() to specify what columns of the data frame will be used for the graphical elements of the plot.
* You must specify what kind of geometric objects or geoms will be displayed: e.g. geom_point(), geom_bar(), geom_boxpot().
* Pretty much anything else that you want to add to your plot is controlled by auxiliary functions, especially those things that have to do with the format, rather than the underlying data.
* The construction of a ggplot is done by adding layers with the + operator.

## Scatterplots

Let's start with a scatterplot of salary and points

```{r}
# scatterplot (option 1)
ggplot(data = dat) +
  geom_point(aes(x = points, y = salary))
```


* ggplot() creates an object of class "ggplot"
* the main input for ggplot() is data which must be a data frame
* then we use the "+" operator to add a layer
* the geometric object (geom) are points: geom_points()
* aes() is used to specify the x and y coordinates, by taking columns points and salary from the data frame

The same scatterplot can also be created with this alternative, and more common use of ggplot()


```{r}
# scatterplot (option 2)
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point()
```


## Label your chunks!

When including code for plots and graphics, we strongly recommend that you create an individual code chunk for each plot, and that you give a label to that chunk. 

When including code for plots and graphics, we strongly recommend that you create an individual code chunk for each plot, and that you give a label to that chunk. 

Note that the code chunk has a label scatterplot1; moreover, the code is exclusively decidated to this plot. Why should you care? Because when "knitr" creates the file of the plot, it will use the chunk label for the graph. So it's better to give meaningful names to those chunks containing graphics.


## Adding color

Say you want to color code the points in terms of position

```{r}
# colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position))
```

Maybe you wan to modify the size of the dots in terms of points3:

```{r}
# sized and colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position, size = points3))
```

To add some transparency effect to the dots, you can use the alpha parameter.

```{r}
# sized and colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position, size = points3), alpha = 0.7)
```

Notice that alpha was specified outside aes(). This is because we are not using any column for the alpha transparency values.

## Your turn:

* Open the ggplot2 cheatsheet

* Use the data frame gsw to make a scatterplot of height and weight.
```{r}
ggplot(data = gsw, aes(x = height, y = weight)) + 
  geom_point()
```

* Find out how to make another scatterplot of height and weight, using geom_text() to display the names of the players.
```{r}
ggplot(data = gsw, aes(x = height, y = weight)) + 
  geom_text(aes(label = player))
```

* Get a scatter plot of height and weight, for ALL the warriors, displaying their names with geom_label().
```{r}
ggplot(data = gsw, aes(x = height, y = weight)) + 
  geom_label(aes(label = player))
```

* Get a density plot of salary (for all NBA players).
```{r}
ggplot(dat, aes(x = salary)) + 
  geom_density(kernel = "gaussian")
```

* Get a histogram of points2 with binwidth of 50 (for all NBA players).
```{r}
ggplot(dat, aes(x = points2)) + 
  geom_histogram(binwidth = 50)
```

* Get a barchart of the position frequencies (for all NBA players).
```{r}
ggplot(dat, aes(x = position)) + 
  geom_bar()
```

* Make a scatterplot of experience and salary of all Centers, and use geom_smooth() to add a regression line.
```{r}
ggplot(filter(dat, position == "C"), aes(x = experience, y = salary)) + 
  geom_point() + 
  geom_smooth(method = lm)
```

* Repeat the same scatterplot of experience and salary of all Centers, but now use geom_smooth() to add a loess line (i.e. smooth line).
```{r}
ggplot(filter(dat, position == "C"), aes(x = experience, y = salary)) + 
  geom_point() + 
  geom_smooth(method = loess)
```

# Faceting

One of the most attractive features of "ggplot2" is the ability to display multiple facets. The idea of facets is to divide a plot into subplots based on the values of one or more categorical (or discrete) variables.

Here's an example. What if you want to get scatterplots of points and salary separated (or grouped) by position? This is where faceting comes handy, and you can use facet_warp() for this purpose:

```{r}
# scatterplot by position
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point() +
  facet_wrap(~ position)
```

The other faceting function is facet_grid(), which allows you to control the layout of the facets (by rows, by columns, etc)

```{r}
# scatterplot by position
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position), alpha = 0.7) +
  facet_grid(~ position) +
  geom_smooth(method = loess)

```

```{r}
# scatterplot by position
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position), alpha = 0.7) +
  facet_grid(position ~ .) +
  geom_smooth(method = loess)
```

## Your turn:

* Make scatterplots of experience and salary faceting by position
```{r}
ggplot(data = dat, aes(x = experience, y = salary)) +
  geom_point(aes(color = experience), alpha = 0.7) +
  facet_grid(position ~ .) 
```

* Make scatterplots of experience and salary faceting by team
```{r}
ggplot(data = dat, aes(x = experience, y = salary)) +
  geom_point() +
  facet_wrap(~ team)
```

* Make density plots of age faceting by team
```{r}
ggplot(data = dat, aes(x = age)) +
  geom_density() +
  facet_wrap(~ team)
```

* Make scatterplots of height and weight faceting by position
```{r}
ggplot(data = dat, aes(x = weight, y = height)) +
  geom_point() +
  facet_wrap(~ position)
```

* Make scatterplots of height and weight, with a 2-dimensional density, geom_density2d(), faceting by position
```{r}
ggplot(data = dat, aes(x = height, y = weight)) +
  geom_point(aes(color = position), alpha = 0.7) +
  facet_grid(~position) +
  geom_density2d()
```

* Make a scatterplot of experience and salary for the Warriors, but this time add a layer with theme_bw() to get a simpler background
```{r}
ggplot(data = filter(dat, team == "GSW"), aes(x = experience, y = salary)) +
  geom_point(aes(color = position), alpha = 0.7) +
  facet_grid(~ position) +
  theme_bw()
```

* Repeat any of the previous plots but now adding a leyer with another theme e.g. theme_minimal(), theme_dark(), theme_classic()
```{r}
ggplot(data = dat, aes(x = age)) +
  geom_density() +
  facet_wrap(~ team) + 
  theme_dark()
```


