---
title: "hw2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "./images/")
```

## 5.1 | Total Shots by Player

```{r}
library(dplyr)

shots_data <- read.csv("/Users/akii_my/Desktop/school/stat133/hw/hw2/data/shots-data.csv")

total_shots_by_player <- shots_data %>%
  group_by(name)%>%
  select(name) %>%
  summarise(total =n()) %>%
  arrange(desc(total))

total_shots_by_player

```

## 5.2 | Effective Shooting Percentage

### Effective Shooting Perentage Player
```{r}
shots_data %>%
  filter(shot_made_flag=="made_shot" | shot_made_flag=="missed_shot") %>%
  group_by(name)%>%
  summarise(total = n()) %>% 
  ungroup() %>% 
  arrange(name) %>%
  data.frame -> data_1

shots_data %>%
  filter(shot_made_flag=="made_shot") %>%
  group_by(name)%>%
  summarise(made = n()) %>% 
  ungroup() %>% 
  arrange(name) %>%
  data.frame -> data_2

shots_data %>%
  filter(shot_made_flag=="missed_shot") %>%
  group_by(name)%>%
  summarise(missed = n()) %>% 
  ungroup() %>% 
  arrange(name) %>%
  data.frame -> data_3
perc_made <- (data_2$made/data_1$total) * 100 
cbind(data_1, select(data_2, made), perc_made) -> effective_shooting
arrange(effective_shooting, desc(perc_made))

```


### Effective Shooting Percentage Player

```{r}
shots_data %>%
  filter((shot_type=="2PT Field Goal" & shot_made_flag=="missed_shot") | (shot_type=="2PT Field Goal" & shot_made_flag=="made_shot")) %>%
  group_by(name)%>%
  summarise(total_2PT = n()) %>% 
  ungroup() %>% 
  arrange(name) %>%
  data.frame -> data_1_2pt


shots_data %>%
  filter((shot_type=="2PT Field Goal" & shot_made_flag=="made_shot")) %>%
  group_by(name)%>%
  summarise(made = n()) %>% 
  ungroup() %>% 
  arrange(name) %>%
  data.frame -> data_2_2pt


perc_made_2pt <- (data_2_2pt$made/data_1_2pt$total) * 100
cbind(data_1_2pt, select(data_2_2pt, made), perc_made_2pt) -> effective_shooting_2pt
arrange(effective_shooting_2pt, desc(perc_made_2pt))

```

### Effective Shooting Percentage Player

```{r}
shots_data %>%
  filter((shot_type=="3PT Field Goal" & shot_made_flag=="missed_shot") | (shot_type=="3PT Field Goal" & shot_made_flag=="made_shot")) %>%
  group_by(name)%>%
  summarise(total_3PT = n()) %>% 
  ungroup() %>% 
  arrange(name) %>%
  data.frame -> data_1_3pt


shots_data %>%
  filter((shot_type=="3PT Field Goal" & shot_made_flag=="made_shot")) %>%
  group_by(name)%>%
  summarise(made = n()) %>% 
  ungroup() %>% 
  arrange(name) %>%
  data.frame -> data_2_3pt


perc_made_3pt <- (data_2_3pt$made/data_1_3pt$total) * 100
cbind(data_1_3pt, select(data_2_3pt, made), perc_made_3pt) -> effective_shooting_3pt
arrange(effective_shooting_3pt, desc(perc_made_3pt))


```


# 6 | Shooting Distance

## 6.1 | Dplyr Table

```{r}
shots_data %>%
  group_by(shot_distance)%>%
  select(shot_distance, shot_made_flag) %>%
  summarise(
    total_shots = n(),
    made_shots = sum(shot_made_flag == "made_shot")) %>%
  mutate(made_shot_prop = made_shots / total_shots) %>%
  select(shot_distance, made_shot_prop)
```


## 6.2 | ggplot
```{r}
library(ggplot2)

shot_distance_table <- data.frame(
      shots_data %>%
      group_by(shot_distance) %>%
      select(shot_distance, shot_made_flag) %>%
      summarise(made_shot_prop = 
                  (sum(shot_made_flag == 'made_shot'))/
                  (sum(shot_made_flag == 'made_shot'|shot_made_flag == 'missed_shot'))
                )
)

ggplot(shot_distance_table, aes(x=shot_distance, y=made_shot_prop)) + geom_point(size=4, shape=20)
```

I can observe that most of the points are below 100 on made_shot_propand the graph itself is a decreasing graph overall. There are outliers such as the first 3 points or the points between 20 and 30 on shotdistance. It can be seen that after 30, the shots made decreases dramatically and hits almost 0 from there. It cannot be said that the shorter the didstance, the more effective the shots becuase the number of shots taken between 20 and 30 are higher in general than the ones from 5 to 20, making the statement false. I guesstimate that after the shot distance of 35, the chance of making successful shot is basically null because after the point, the number of successful shots become 0.

```{r}
shot_distance_table$shot_distance[shot_distance_table$made_shot_prop >= 0.50]


outlier_shot <- shot_distance_table %>%
  filter(shot_distance > 40) %>%
  filter(made_shot_prop == 1)

outlier_shot$shot_distance
```

The distances that tend to have a percentage of 50% or more is 0, 1, 12, 16, 18, 21, 30, and 51.
There is also an atypical (outlier) shot successfully made from 51 feet. 


# Total number of shots by minutes of occurrence
```{r}
shots_in_minute <- shots_data %>%
  group_by(name, minute) %>%
  summarise(number_shots = n())
shots_in_minute


shots_in_mn_plot <- ggplot(shots_in_minute,
                           aes(x = minute, y = number_shots)) +
  geom_vline(xintercept = c(12, 24, 36, 48), color = 'gray80') +
  scale_x_continuous(breaks = c(1, 12, 24, 36, 48)) +
  ylab('total number of shots') +
  geom_rect(aes(xmin = 0, xmax = 12, ymin = -Inf, ymax = Inf),
            fill = "gray95") +
  geom_rect(aes(xmin = 24, xmax = 36, ymin = -Inf, ymax = Inf),
            fill = "gray95") +
  theme_minimal() +
  geom_path(color = '#498dfc55', size = 1) +
  geom_point(color = "#498dfc", size = 2) +
  facet_wrap(~ name) +
  ggtitle('Total number of shots (by minute of occurrence)')

shots_in_mn_plot 
```


