---
title: "Lab 4: Data Frame Basics"
author: "Amy Kim"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abalone Data Set

The first data set that you will working with is the Abalone Data Set that is part of the UCI Machine Learning Repository

The location of the data file is:

[http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data](http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data)

The location of the data dictionary (description of the data) is:

[http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names](http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names)

Look at both the dataset file, and the file with its description, and answer the following questions:

*What's the character delimiter?

The character delimiter is ",".

*Is there a row for column names?

No. 

*Are there any missing values? If so, how are they codified?

No missing values. 


*What is the data type of each column?

Name | Data Type	
-----|-------------------
Sex	|	nominal	
Length |	continuous
Diameter |	continuous
Height	|	continuous
Whole weight |	continuous
Shucked weight |	continuous
Viscera weight	| continuous	
Shell weight	| continuous
Rings	|	integer	

One basic way to read this file in R is by passing the url location of the file directly to any of the read.table() functions:

```{r}
url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data"
abalone <- read.table(url, sep = ",")
```

## Getting a Local Copy of the Data

My suggestion when reading datasets from the Web, is to always try to get a local copy of the data file in your machine (as long as you have enough free space to save it in your computer). To do this, you can use the function download.file() and specify the url address, and the name of the file that will be created in your computer. For instance, to save the abalone data file in your working directory, type the following commands directly on the R console:

# Some Bash Commands

Before describing some of the reading-table functions in R, let's practice some basic bash commands to inspect the downloaded data file.

* For those of you using Gitbash, go to your browser and open a tab to get the Linux Man Pages Online available in the website: http://man.he.net/

* Open the terminal (e.g. Mac terminal or Gitbash) and change directories to the folder that contains the downloaded file abalone.data.

* Use the file command to know what type of file is abalone.data. 
```{r}
#CSV text
```


* Use the word count command wc to obtain information about: 1) newline count, 2) word count, and 3) byte count, of the abalone.data file. 
```{r}
# 1) 4177 
# 2) 4177
# 3) 191873
```

* See the man documentation of wc and learn what option you should use to obtain only the number of lines in abalone.data.
```{r}
#wc -l 
# 4177
```

* Use head to take a peek at the first lines (10 lines by default) of abalone.data
```{r}
#M,0.455,0.365,0.095,0.514,0.2245,0.101,0.15,15
#M,0.35,0.265,0.09,0.2255,0.0995,0.0485,0.07,7
#F,0.53,0.42,0.135,0.677,0.2565,0.1415,0.21,9
#M,0.44,0.365,0.125,0.516,0.2155,0.114,0.155,10
#I,0.33,0.255,0.08,0.205,0.0895,0.0395,0.055,7
#I,0.425,0.3,0.095,0.3515,0.141,0.0775,0.12,8
#F,0.53,0.415,0.15,0.7775,0.237,0.1415,0.33,20
#F,0.545,0.425,0.125,0.768,0.294,0.1495,0.26,16
#M,0.475,0.37,0.125,0.5095,0.2165,0.1125,0.165,9
#F,0.55,0.44,0.15,0.8945,0.3145,0.151,0.32,19
```

* See the man documentation of head and learn what option you should use to display only the first 5 files in abalone.data.
```{r}
#head -5 abalone.data
#M,0.455,0.365,0.095,0.514,0.2245,0.101,0.15,15
#M,0.35,0.265,0.09,0.2255,0.0995,0.0485,0.07,7
#F,0.53,0.42,0.135,0.677,0.2565,0.1415,0.21,9
#M,0.44,0.365,0.125,0.516,0.2155,0.114,0.155,10
#I,0.33,0.255,0.08,0.205,0.0895,0.0395,0.055,7
```

* Use tail to take a peek at the last lines (10 lines by default) of abalone.data
```{r}
#tail abalone.data
#M,0.5,0.38,0.125,0.577,0.269,0.1265,0.1535,9
#F,0.515,0.4,0.125,0.615,0.2865,0.123,0.1765,8
#M,0.52,0.385,0.165,0.791,0.375,0.18,0.1815,10
#M,0.55,0.43,0.13,0.8395,0.3155,0.1955,0.2405,10
#M,0.56,0.43,0.155,0.8675,0.4,0.172,0.229,8
#F,0.565,0.45,0.165,0.887,0.37,0.239,0.249,11
#M,0.59,0.44,0.135,0.966,0.439,0.2145,0.2605,10
#M,0.6,0.475,0.205,1.176,0.5255,0.2875,0.308,9
#F,0.625,0.485,0.15,1.0945,0.531,0.261,0.296,10
#M,0.71,0.555,0.195,1.9485,0.9455,0.3765,0.495,12
```

* See the man documentation of tail and learn what option you should use to display only the last 3 files in abalone.data.
```{r}
# tail -3 abalone data
#M,0.6,0.475,0.205,1.176,0.5255,0.2875,0.308,9
#F,0.625,0.485,0.15,1.0945,0.531,0.261,0.296,10
#M,0.71,0.555,0.195,1.9485,0.9455,0.3765,0.495,12
```


* Use the less command to look at the contents of abalone.data (this command opens a paginator so you can move up and down the contents of the file).

# Basic Importing

Now that you have a local copy of the dataset, you can read it in R with read.table() like so:

```{r}
# reading data from your working directory
abalone <- read.table("abalone.data", sep = ",")
```

Once you read a data table, you may want to start looking at its contents, usually taking a peek at a few rows. This can be done with head() and/or with tail():

```{r}
# take a peek of first rows
head(abalone)

# take a peek of last rows
tail(abalone)
```

Likewsie, you may also want to examine how R has decided to take care of the storage details (what data type is used for each column?). Use the function str() to check the structure of the data frame:
```{r}
# check data frame's structure
str(abalone, vec.len = 1)
```

## Detailed information about the columns

So far we have been able to read the data file in R. But we are missing a few things. First, we don't have names for the columns. Second, it would be nice if we could specify the data types of each column instead of letting R guess how to handle each data type.

According to the description of the Abalone data set, we can assign the following data types to each of the columns as:

Name | Data Type
-----|-------------
Sex  | character
Length | continuous
Diameter | continuous
Height | continuous
Whole weight | continuous
Shucked weight | continuous
Viscera weight | continuous
Shell weight | continuous
Rings | integer


Let's create a vector of columns names, and another vector of data types:

```{r}
# vector of column names
column_names <- c(
    'sex',
    'length',
    'diameter',
    'height',
    'whole_weight',
    'shucked_weight',
    'viscera_weight',
    'shell_weight',
    'rings'
)

# vector of data types (for each column)
column_types <- c(
    'character',
    'real',
    'real',
    'real',
    'real',
    'real',
    'real',
    'real',
    'integer'   
)

```

Optionally, we could also specify a type "factor" for the variable sex since this is supposed to be in nominal scale (i.e. it is a categorical variable). Also note that the variable rings is supposed to be integers, therefore we can choose an integer vector for this column.

Now we can re-read the table in a more complete (and usually more efficient) way:

```{r}
abalone <- read.table(
    'abalone.data',
    col.names = column_names,
    colClasses = column_types,
    sep = ","
)

# check its structure again
str(abalone, vec.len = 1)
```

## Your Turn 

* Read the Abalone data with the read.csv() function.
* Use the inputs col.names and colClasses to specify column names and their data types.
* Look at the data description in the following link: http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names and confirm the following statistics:

```{r}
csv.abalone <- read.csv('abalone.data', col.names = column_names, 
         colClasses = column_types,
         sep = ",")

summary <- summary(csv.abalone, digits = max(3))
summary[c(1,3,4,6), -c(1)]

```

# Import Abalone data with read_csv()

In addition to the built-in functions for importing tables in R, there is also the set of functions from the R package "readr":

* read_delim()
* read_csv()
* read_tsv()
* read_csv2()
* read_fwf()
* read_table()

Take a look at this post for some examples:https://blog.rstudio.com/2015/04/09/readr-0-1-0/

* Use read_csv() to import the abalone data set.
* Use the argument col_names to specify the column names.
* Learn how to use the argument col_types to specify the data type for each column.

```{r}
library(readr)

spec_csv("abalone.data")


csv <- read_csv("abalone.data", col_names = column_names, 
                col_types = list(
                  sex = col_character(), 
                  length = col_double(),
                  diameter = col_double(),
                  height = col_double(),
                  whole_weight = col_double(),
                  shucked_weight = col_double(),
                  viscera_weight = col_double(),
                  shell_weight = col_double(),
                  rings = col_double()
                ))

```

# Pittsburgh Bridges Data Set

This data set is part of the UCI Machine Learning Repository:

[http://archive.ics.uci.edu/ml/datasets/Pittsburgh+Bridges](http://archive.ics.uci.edu/ml/datasets/Pittsburgh+Bridges)

The data Description is here:

[http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.names](http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.names)

The Data file is here:

[http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1](http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1)

Read the description, and take a look at the data set:

* Are there column names? 

Yes. The column names are IDENTIF(identification), RIVER, LOCATION, ERECTED, PURPOSE, LENGTH, LANES, CLEAR-G, T-OR-D, MATERIAL, SPAN n, REL-L, TYPE n 

* What is the field separator?

The field separator is ",". 

* Are there any missing values?

Yes. there are missing values. 

* What is the character for missing values (if any)?

The character for missing values is "?"

* What is the data type of each variable (i.e. column)?

Name | Type
-----|-------------
RIVER | nominal 
LOCATION | nominal 
ERECTED | conitnuous/integer , nominal 
PURPOSE | nominal 
LENGTH | conitnuous/integer , nominal 
LANES | conitnuous/integer , nominal 
CLEAR-G | nominal 
T-OR-D | nominal 
MATERIAL | nominal 
SPAN | nominal 
REL-L | nominal 
TYPE | nominal 

* Download a copy of the data to your computer (use download.file()) and save it in a file named bridges.data

## Reading the Data

* Create a vector of column names 
```{r}
column_names <- c(
  'identification',
  'river',
  'location',
  'erected',
  'purpose',
  'length',
  'lanes',
  'clear-g',
  't-or-d',
  'material',
  'span',
  'rel-l',
  'type') 

```

* Create a vector of column types
```{r}
column_types <- c(
  'character', 
  'character',
  'double',
  'double',
  'character',
  'double',
  'double',
  'character',
  'character',
  'character',
  'character',
  'character',
  'character'
)
```

* Use the function read.table() to import the data. Name it bridges1.
```{r}
bridges1 <- read.table('bridges.data',
    col.names = column_names,
    sep = ",")

bridges1[bridges1 == "?"] <- NA

```

* Use the function read.csv() to import the data. Name it bridges2.
```{r}
bridges2 <- read.csv('bridges.data', stringsAsFactors = FALSE,
                     sep = ",",
                     na.strings = "NULL",
                     col.names = column_names)

bridges2[bridges2 == "?"] <- NA
```


## Basic Inspection 

Use functions to start examining the bridges1 data frame:

* str()
```{r}
str(bridges1)
```

* summary()
```{r}
summary(bridges1)
```

* head() and tail()
```{r}
head(bridges1)
tail(bridges1)
```

* dim()
```{r}
dim(bridges1)
```

* names()
```{r}
names(bridges1)
```

* colnames()
```{r}
colnames(bridges1)
```

* nrow()
```{r}
nrow(bridges1)
```

* ncol()
```{r}
ncol(bridges1)
```


## Optional: Want to do more?

Write R code to find:

* Year of the oldest erected bridge
```{r}
min(bridges1$erected)
```

* Year of the most recent erected bridge
```{r}
max(bridges1$erected)
```

* Frequency of bridges by purpose
```{r}
library(plyr)
count(bridges1, 'purpose')
```

* Frequency of materials
```{r}
count(bridges1, 'material')
```

* Average length of the bridges
```{r}
bridges1$length <- as.numeric(bridges1$length)

mean(na.omit(bridges1$length))
```


# Creating Data Frames

Here's a table with the starting lineup of the Golden State Warriors:

Player | Position | Salary | Points | PPG | Rookie
-------|----------|--------|--------|-----|--------
Thompson | SG | 16663575 | 1742 | 22.3 | FALSE
Curry | PG | 12112359 | 1999 | 25.3 | FALSE
Green | PF | 15330435 | 776 | 10.2 | FALSE
Durant | SF | 26540100 | 1555 | 25.1 | FALSE
Pachulia | C | 2898000 |426 | 6.1 | FALSE


* Start by creating vectors for each of the columns.
```{r}
player <- c("Thompson", "Curry", "Green", "Durant", "Pachulia")
position <- c("SG", "PG", "PF", "SF", "C")
salary <- c(16663575, 12112359, 15330435, 26540100, 2898000)
points <- c(1742, 1999, 775, 1555, 426)
ppg <- c(22.3, 25.3, 10.2, 25.1, 6.1)
rookie <- c(FALSE, FALSE, FALSE, FALSE, FALSE)
```

* Use the vectors to create a first data frame with data.frame().
```{r}
df <- data.frame(cbind(player, position, salary, points, ppg, rookie))
```


* Check that this data frame is of class "data.frame", and that it is also a list.
```{r}
class(df)
```


* Create another data frame by first starting with a list(), and then passing the list to data.frame().
```{r}
col_names <- c("player", "position", "salary", "points", "ppg", "rookie")
as.data.frame(list(player, position, salary, points, ppg, rookie), col.names = col_names)
```


* What would you do to obtain a data frame such that when you check its structure str() the variables are:

- Player as character
- Position as factor
- Salary as numeric or real (ignore the commas)
- Points as integer
- PPG as numeric or real
- Rookie as logical

```{r}
df$position <- as.factor(df$position)
df$salary <- as.numeric(df$salary)
df$points <- as.integer(df$points)
df$ppg <- as.numeric(df$ppg)
df$rookie <- as.logical(df$rookie)
str(df)
```

* Find out how to use the column binding function cbind() to create a tabular object with the vectors created in step 1 (inspect what class of object is obtained with cbind()).

```{r}
bind <- cbind(player, position, salary, points, ppg, rookie)

class(bind)
```


* How could you convert the object in the previous step into a data frame?
```{r}
as.data.frame(bind)
```


# NBA Players Data

Now that you've seen some of the most basic operations to manipulate data frames, let's apply them on a data set about NBA players. The corresponding data file is nba2017-players.csv, located in the data/ folder of the course github repository. This file contains 15 variables measured on 441 players.

First download a copy of the csv file to your computer.

To import the data in R you can use the function read.csv():
```{r}
csv <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/data/nba2017-players.csv"
download.file(url = csv, destfile = 'nba2017-players.csv')

dat <- read.csv('nba2017-players.csv', stringsAsFactors = FALSE)
```

Notice that I'm specifying the argument stringsAsFactors = FALSE to avoid the conversion of characters into R factors.

All the default reading table functions generate a data frame. Typically, everytime I read a new data set which I'm not familiar with, or a data set that I haven't worked on in a long time, I always like to call a couple of functions to inspect its contents:

* dim()
* head()
* tail()
* str()
* summary()

A first check-up is to examine the dimensions of the data frame with dim():
```{r}
# dimensions (# of rows, # of columns)
dim(dat)
```

If you know in advanced how many rows and columns are in the data table, this is a good way to make sure that R was able to read all the records.

Then, depending on the size of the data, you may want to take a peek at its contents with head() or tail(), just to get an idea of what the data looks like:

```{r}
# display first few rows
head(dat)
```

For a more detailed description of how R is treating the data type in each column, you should use the structure function str().

```{r}
# check the structure
str(dat, vec.len = 1)
```

This function str() displays the dimensions of the data frame, and then a list with the name of all the variables, and their data types (e.g. chr character, num real, etc). The argument vec.len = 1 indicates that just the first element in each column should be displayed.

When working with data frames, remember to always take some time inspecting the contents, and checking how R is handling the data types. It is in these early stages of data exploration that you can catch potential issues in order to avoid disastrous consequences or bugs in subsequent stages.

Your turn:
Use bracket notation, the dollar operator, as well as concepts of logical subsetting and indexing to:

Display the last 5 rows of the data.
```{r}
tail(dat)
```

Display those rows associated to players having height less than 70 inches tall.
```{r}
dat[dat$height < 70, ]
```

Of those players that are centers (position C), display their names and salaries.
```{r}
center <- dat[dat$position == "C", ]
center[c("player", "salary")]
```


Create a data frame durant with Kevin Durant's information (i.e. row).
```{r}
which(dat$player == "Kevin Durant")
durant <- as.data.frame(dat[228,])
durant
```

Create a data frame ucla with the data of players from college UCLA ("University of California, Los Angeles").
```{r}
ucla <- as.data.frame(dat[c(which(dat$college == "University of California, Los Angeles")),])
```

Create a data frame rookies with those players with 0 years of experience.
```{r}
rookies <- as.data.frame(dat[which(dat$experience == 0), ])
```

Create a data frame rookie_centers with the data of Center rookie players.
```{r}
rookie_centers <- as.data.frame(dat[which(dat$experience == 0 & dat$position == "C"),])
```

Create a data frame top_players for players with more than 50 games and more than 100 minutes played.
```{r}
top_players <- as.data.frame(dat[which(dat$games > 50 & dat$minutes > 100),])
```

What's the largest height value?
```{r}
max(dat$height)
```

What's the minimum height value?
```{r}
min(dat$height)
```

What's the overall average height?
```{r}
mean(dat$height, na.rm = TRUE)
```

Who is the tallest player?
```{r}
dat$player[which.max(dat$height)]
```

Who is the shortest player?
```{r}
dat$player[which.min(dat$height)]
```

Which are the unique teams?
```{r}
count(dat$team)[,1]
```

How many different teams?
```{r}
nrow(count(dat$team))
```

Who is the oldest player?
```{r}
dat$player[which.max(dat$age)]
```

What is the median salary of all players?
```{r}
median(dat$salary)
```

What is the median salary of the players with 10 years of experience or more?
```{r}
experience <- dat[dat$experience > 10 | dat$experience == 10, ]
median(experience$salary)
```

What is the median salary of Shooting Guards (SG) and Point Guards (PG)?
```{r}
sgpg <- dat[dat$position == "SG" | dat$position == "PG", ]
median(sgpg$salary)
```

What is the median salary of Power Forwards (PF), 29 years or older, and 74 inches tall or less?
```{r}
dat[dat$position == "PF" & dat$age > 29 & dat$height < 74 | dat$position == "PF" & dat$age > 29 & dat$height == 74 | dat$position == "PF" & dat$age == 29 & dat$height == 74 | dat$position == "PF" & dat$age == 29 & dat$height < 74,]

```

No player matches the criteria 

How many players scored 4 points or less?
```{r}
length(dat[dat$points == 4 | dat$points < 4, ])
```

Who are those players who scored 4 points or less?
```{r}
dat[dat$points == 4 | dat$points < 4, ][,1]
```

Who is the player with 0 points?
```{r}
dat[dat$points == 0,][,1]
```

How many players are from "University of California, Berkeley"?
```{r}
length(dat[dat$college == "University of California, Berkeley"])
```

Are there any players from "University of Notre Dame"? If so how many and who are they?
```{r}
dat[dat$college == "University of Notre Dame", ][,1]

length(dat[dat$college == "University of Notre Dame", ][,1])
```

Are there any players with weight greater than 260 pounds? If so how many and who are they?
```{r}
dat[dat$weight > 260, ][,1]
length(dat[dat$weight > 260, ][,1])
```

Who is the player with the maximum rate of points per minute?
```{r}
dat[ which.max(dat$points/60),][,1]
```

Who is the player with the maximum rate of three-points per minute?
```{r}
dat[ which.max(dat$points3/60),][,1]
```

Who is the player with the maximum rate of two-points per minute?
```{r}
dat[ which.max(dat$points2/60),][,1]
```

Who is the player with the maximum rate of one-points (free-throws) per minute?
```{r}
dat[ which.max(dat$points1/60),][,1]
```

Create a data frame gsw with the name, height, weight of Golden State Warriors (GSW)
```{r}
gsw <- dat[,c("player","height", "weight")]
```

Display the data in gsw sorted by height in increasing order (hint: see ?sort and ?order)
```{r}
gsw[order(gsw$height),]
```

Display the data in gsw by weight in decreasing order (hint: see ?sort and ?order)
```{r}
gsw[order(gsw$height, decreasing = TRUE),]
```

Display the player name, team, and salary, of the top 5 highest-paid players (hint: see ?sort and ?order)
```{r}
head(dat[order(dat$salary, decreasing = TRUE),])[-5,c("player","team", "salary")]
```

Display the player name, team, and points3, of the top 10 three-point players (hint: see ?sort and ?order)
```{r}
dat[order(dat$points3, decreasing = TRUE),][1:10,c("player","team","points3")]
```


# Group By 

Group-by operations are very common in data analytics. Without dedicated functions, these operations tend to be very hard (labor intensive).

So what functions can you use in R to perform group by operations? In base R, the main function for group-by operations is aggregate().

Here's an example using aggregate() to get the median salary, grouped by team:

```{r}
aggregate(dat$salary, by = list(dat$team), FUN = median)
```

The same example above can also be obtained with aggreagte() using formula notation like this:

```{r}
aggregate(salary ~ team, data = dat, FUN = median)
```

Here's another example using aggregate() to get the average height and average weight, grouped by position:

```{r}
aggregate(dat[ ,c('height', 'weight')], by = list(dat$position), FUN = mean)
```

The same example above can also be obtained with aggreagte() using formula notation like this:

```{r}
aggregate(. ~ position, data = dat[ ,c('position', 'height', 'weight')], 
          FUN = mean)
```

## Your turn

Create a data frame with the average height, average weight, and average age, grouped by position
```{r}
aggregate(. ~ position, data = dat[ ,c('position', 'height', 'weight','age')], 
          FUN = mean)
```

Create a data frame with the average height, average weight, and average age, grouped by team
```{r}
aggregate(. ~ team, data = dat[ ,c('team', 'height', 'weight','age')], 
          FUN = mean)
```

Create a data frame with the average height, average weight, and average age, grouped by team and position.
```{r}
aggregate(dat[ ,c('height', 'weight','age')], by = list(dat$position, dat$team), FUN = mean)
```

Difficult: Create a data frame with the minimum salary, median salary, mean salary, and maximum salary, grouped by team and position.
```{r}
aggregate(dat[ ,c('salary')], by = list(dat$position, dat$team), FUN = function(x) c("min" =min(x), "median" = median(x), "mean" = mean(x), "max" = max(x)))

```





