---
title: "Introduction to dieroller"
output: github_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dieroller)
```


## About

The package `"dieroller"` is a minimal implementation for simulating rolling
a die a given number of times, and to visualize the relative frequencies
in such a series of rolls


## Creating a `die`

The first step is to create a `"die"` object with the function `die()`:

```{r}
mydie <- die()
mydie
```

By default, `die()` creates a _fair_ standard die with six sides numbered 1, 2, 3, 4, 5, and 5, and with probability values for each side of 1/6.

You can change the defatuls with arguments `sides` and `prob`:

```{r}
loaded <- die(prob = c(0.1, 0.1, 0.2, 0.2, 0.35, 0.05))
loaded
```


## Rolling a die

Once you have defined a `"die"` object, you can use the function `roll()` to
_roll_ a die a given number of `times`. The output will be an object of class 
`"roll"`, which will contain the vector of `rolls`:

```{r}
rolls100 <- roll(mydie, times = 100)
rolls100
```

The `summary()` function gives you a nice table of frequencies:

```{r}
summary(rolls100)
```


## Plot of relative frequencies

You can use the function `plot()` to create a graohic with the relative 
frequencies of a series of coin tosses:

```{r, fig.show='hold'}
plot(rolls100)
```